---
title: "Topics in Microeconomics - Exercise 5"
output: pdf_document
---

```{r setup, include=FALSE}
# Local env setup
knitr::opts_chunk$set(echo = TRUE)
library(haven)
library(ggplot2)
library(tidyverse)
library(ivreg)
library(rdrobust)
library(sandwich)
library(lmtest)
library(kableExtra)
```

## Exercise 5

These solutions were written by:

1. Lothar Mateo
2. Francisco Velázquez
3. Arturo Soberón

###  Political Selection and Economic Policy

#### a)
_Why is it not obvious how politician quality would map into policy outcomes?_

The relationship between quality and policy outcomes is hard to quantify because we cannot perfectly observe
a politician's quality. For instance, the traits that voters associate to political quality may be correlated
with other factors that could influence a politician's ideology. As an example, assuming that wealthy individuals
prefer right-wing policies, then a candidate regarded as a high-quality politician due to their level of education
may also favor right-wing policies because their education is related to their level of wealth.

In general, the traits that voters associate with a politician's quality may be related to the kind of policies
implemented by those politicians.

#### b)
_How does the author motivate the focus on public spending and fiscal policy? What other outcomes would be
interesting to analyze?_

The author studies these two outcomes because of their potential trade-off. In practice, left-wing politicians tend to
favor public spending, while right-wing politician tend to favor fiscal health. In theory, a low-quality politician
will implement their favorite policies at the expense of other objectives. For example, liberals would increase public
spending and disregard fiscal health, while conservatives would cut public spending to improve fiscal health.

One of the main results of the article is that, in contrast to low-quality politicians, high-quality candidates are
capable of implementing their preferred policies without compromising policies at the opposite end of the ideological
spectrum.

#### c)
_Suppose that a researcher would just regress a policy outcome on the quality of politicians. What kind of problems
would this kind of regression involve?_

The model would not account for omitted variable bias due to unobserved voter preferences or reverse causality due to
underlying relationships, such as politicians' traits being related to their preferred policies. Going back
to the example of a highly educated and wealthy politician, they may be inclined to reduce public spending to improve
fiscal health. This reduction in public goods provision would affect certain populations and the model would yield
a negative relationship between perceived quality and policy outcome.

_To deal with the endogeneity issues, the paper uses an identification strategy like the close elections regression
discontinuity designs that we have already seen in this class. Why is it difficult to conduct a conventional RDD in the
context of the study? Briefly summarize the idea of the identification strategy that the paper uses._

The author compares close elections that involve two different types of politicians from the same party. Meriläinen
calculates the difference between the seat share of a political type against its expected seat share and uses it as an
instrumental variable. This methodology allows the author to regress policy outcome $Y$ in municipality $m$ at time $t$
on the instrumented share of competent politicians.

When constructing the instrumental variables, the first step involves comparing candidates involved in close elections.
The author defines a cutoff value as the average of the within-party vote shares from the last-elected candidate and
the first-non-elected candidate. All candidates with distance from election within $\varepsilon$ are considered to be
involved in _close elections_.

Running a common RDD is not straightforward because there is no well-defined running variable because close elections
can happen withing multiple parties at the same time. The author quotes an extreme example where two similar-quality
politicians get different election outcomes in different parties. Bot of them are within a similar margin from election,
yet one of the was elected in party A, while the other was not in party B.

#### d)
_Why might education be an imperfect indicator of politician quality? What alternative measures does the paper use?
What are the advantages and disadvantages of these? If there were no data limitations, what kind of measure do you
think would be an ideal measure of politician quality and why?_

Education could be related to a politician's ideology due to unobserved factors that shape politicians' ideologies and
are therefore related to political outcomes. For example, education could be related to social status, and thus, a high
level of education could cause the politician to be conservative through this channel.

The author uses data on candidates to regress their income on observable factors, such as age, sector and educations.
The estimated error term from this regression is used as a measure for the unobservable traits that impact income and
therefore candidate quality.

If there were no limitations, a good measure for candidate quality would be their leverage ability. An observable
measure of this variable would be how many policies a given politician has been able to pass in previous years when
facing opposition from other politicians.

#### e)
_How do different politician qualities affect economic policies? What kind of explanation does the author offer for
the fact that different qualities may matter for policy in different ways?_

In the first part of the paper, Meriläinen shows that high-quality candidates are able to increase spending without
compromising the municipality's financial health, or likewise, improve financial health without reducing public goods
provision. In the second part, Meriläinen uses survey data on candidates' ideologies to suggest that more-educated
politicians tend to favor left wing policies (government growth), while high-income and experienced politicians tend to
favor fiscal health.

\newpage
### 2. The Political Resource Curse

#### a)
_What are these predictions made by the theoretical model this paper is based on?_

1. Voters are less likely to notice if the incumbent steals resources when the budget size increases.
2. A larger budget size decreases the average ability of candidates challenging the incumbent because political rents
are more valuable to low-ability candidates.
3. An incumbent facing lower-quality competitors can misbehave more without hurting their reelection prospects.
4. An increase of government resources increases the reelection probability of the incumbent.
5. Electoral punishment is weaker when an incumbent receives a larger budget.
6. Higher transfers increase corruption more if the opponents have lower levels of education.

\newpage
#### b)
_RDD plots_

All of the following plots suggest that receiving additional funds from the government increases the broad corruption
measure in municipalities around the population cutoff.

_(i) Linear polynomial_
```{r}
# Read data sets
d1 <- read_dta('../dat/AER_smallsample.dta')
d2 <- read_dta('../dat/AER_largesample.dta')

# Data for plots
d <- d1 %>% 
  filter(abs(popnorm) <= 3500) %>% 
  select(broad, popnorm)

# Linear model
rdplot(y = d$broad, x = d$popnorm, c = 0, p = 1,
       title = 'Linear RDD',
       x.label = 'Normalized Population',
       y.label = 'Broad corruption',
       col.dots = 'skyblue',
       col.lines = 'coral')
```
The linear model suggests that broad corruption increases after the cutoff. However, the residuals do not seem to be
normally distributed, especially on the right side of the plot. I personally think the linear model is not a good fit.

Based on the plot, the Local Average Treatment Effect (LATE) is an increase in broad corruption of around 12
percentage points.

_(ii) Quadratic polynomial_
```{r}
rdplot(y = d$broad, x = d$popnorm, c = 0, p = 2,
       title = 'Quadratic RDD',
       x.label = 'Normalized Population',
       y.label = 'Broad corruption',
       col.dots = 'skyblue',
       col.lines = 'coral')
```
The fit of the quadratic model is adequate on the left-hand side of the plot. However, the effect seems exaggerated
do to the pronounced curve of the curve on the right. This happens because of the dispersion of the corruption measure
around `popnorm = 2000` (it ranges between 0.6 and 0.9).

Based on the plot, the LATE is an increase in broad corruption of around 15 percentage points.

\newpage
_(iii) Cubic polynomial_
```{r}
# Generate bins
rdplot(y = d$broad, x = d$popnorm, c = 0, p = 3,
       title = 'Cubic RDD',
       x.label = 'Normalized Population',
       y.label = 'Broad corruption',
       col.dots = 'skyblue',
       col.lines = 'coral')
```
The left-hand side of the plot now smoothly follows the points in the graph. The right-hand side of the plot did not
change substantially from the quadratic fit, as there is no point of inflection. As with all previous models, the
corruption measure seems to increase after the cutoff.

In accordance to the quadratic model, the ATT is an increase in broad corruption of around 15 percentage points.

\newpage
#### c)
_Replicate Table 9_

_(i) Education of opponents and reelection of incumbent
```{r}
# Filter data according to seven thresholds
d <- d2 %>% 
  filter(pop <= 50940)

# Dependent variables
Y <- c('opp_college', 'opp_yschool', 'reele_inc')

# Columns to dummies
d$term <- as.factor(d$term)
d$regions <- as.factor(d$regions)

# General formula
f <- '~ pop + pop_2 + pop_3 + term + regions + fpm | pop + pop_2 + pop_3 + term + regions + fpm_hat'

# Initialize empty dataframe to store coefficients in
res <- matrix(nrow = 2, ncol = 0) %>% as.data.frame()

# For loop for all variables
for (y in Y) {
  # Model
  m <- ivreg(
    formula = paste(y, f),
    data = d
  )
  
  # Clustered SE
  m <- coeftest(x = m, vcov = vcovCL, cluster = ~id_city)
  
  # Pull effect of fpm
  m <- m['fpm',]
  
  # Append to dataframe
  res <- res %>% 
    mutate(
      !!y := c(m['Estimate'], m['Std. Error']) %>% unname()
    )
}

# Rename columns
names(res) <- c('College','Years of schooling','Incumbent reelection')

# Results
res %>% 
  round(3) %>% 
  kbl()
```

_(ii) Interpretation_
The regression results indicate that an increase in FPM leads to a decrease in the share of opponents with a college
degree, a reduction in the average years of schooling of opponents', and an increase in the probability that the
incumbent will be reelected.

With all other values held constant, a one-unit increase in FPM causes:
1. A 0.8 percentage point decrease in the share of opponents with a college degree.
2. A decrease in the average years of schooling of opponents of 0.077 years.
3. An increase in the probability of reelection of 1.2 percentage points.

These results agree with those presented in Figure 3 because, according to the RDDs, having a slightly larger 
population count (and hence getting more funds) has effects in the same directions as the IV regressions. The main
difference is that Figure 3 uses RDD with population as a running variable, whereas Table 9 uses an IV approach
to measure the same effect (receiving more funds).

_(iii) Removing controls_
We consider that the controls are necessary because of the inclusion of population. If we were to remove population,
then the models would be interpreted as an increase in FPM without regard of a municipality's population count.
By construction of the Brazilian system, municipalities with larger populations receive larger cuts of the state's
budget, and hence, the marginal effects of an increase in population count should differ depending on the bracket.


#### d) _Histogram of the running variable_

The distribution of the running variable is skewed to the right. Normally, we would be worried that municipalities
can strategically manipulate their reported population counts. However, since greater populations translate to greater
FPM coefficients, then it seems that this is the actual distribution at the cutoff.

We think this distribution does not invalidate the RDD because it would not make sense for municipalities to
under-report their population counts because that would translate to receiving less federal funds.

Plotting the distribution of the running variable is a good strategy to determine if the underlying population
is strategically manipulating its reports in order to become eligible to receive the funds from the policy.

In addition to studying the distribution of the running variable, we can run RDDs on the controls with the intent to
show that the table is balanced around the cutoff. That is, all variables except for the treatment are balanced, which
is evidence that the assignment of the treatment is as good as random around the cutoff.
```{r}
ggplot(data = d2, mapping = aes(x = popnorm)) +
  geom_histogram(bins = 30, fill = 'skyblue') +
  geom_vline(aes(xintercept = 0), colour = 'black', linetype = 'dashed') +
  ggtitle('Histogram of Population') +
  xlab('Population') +
  ylab('Count by bin') +
  theme_light()
```

#### e)
_(i) What potential problems could there be with studies that exploit this kind of RDD set-ups?_
According to Cattaneo et al., when there are multiple cutoffs in a policy, using an RDD to analyze the local
treatment effect for all buckets in a single model may not be able to capture all the information available in the
multiple cutoffs. In particular, compound treatments and sorting are what causes normal RDDs to not be able to fully
retrieve the causal effect of a policy on an outcome.

_(ii) How to deal with multiple population thresholds?_
In order to deal with compound policies, researches should look for other settings where the confounding policies
change but the policy of interest does not. This can be done with a difference-in-discontinuity design.

Sorting creates bias in RDD estimates because the distribution of covariates differs on both sides of the cutoff.
Therefore, the solution proposed by Eggers et al. is to measure these covariates and their relationship to the outcome
of interest at the threshold. In other words, researchers must control for these covariates in order to make the
RDD look like a typical randomized experiment.

_(iii) Can you think of any population-based policies that would apply in the Mexican context?_
Mexican families are given cash transfers if their children are enrolled in school, are not malnourished and receive
regular check-ups. The running variable could be households' Capital Income, as families become eligible for these
transfers if they fall below the minimum well-being line.
